---
title: "Final Data Engineering Prep"
subtitle: "Re-examining Singapore's Declining Fertility"
author: "Team 11"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This notebook prepares the first working version of the data engineering pipeline for the final submission.

It uses the local datasets already available in the project folder and reshapes them into visualization-ready tables. Each section can be run one by one in RStudio.

## 1. Setup

```{r}
library(tidyverse)
```

## 2. File paths

```{r}
data_dir <- file.path("proposal_files", "datasets")

fertility_file <- file.path(data_dir, "BirthsAndFertilityRatesAnnual.csv")
singlehood_file <- file.path(data_dir, "M810581.csv")
lfpr_file <- file.path(data_dir, "M182201.csv")

tibble(
  dataset = c("Births and fertility", "Singlehood", "Female LFPR"),
  file_path = c(fertility_file, singlehood_file, lfpr_file),
  file_exists = file.exists(c(fertility_file, singlehood_file, lfpr_file))
)
```

## 3. Helper functions


```{r}
read_singstat_wide <- function(path) {
  lines <- readLines(path, warn = FALSE)
  header_index <- which(stringr::str_detect(lines, "^Data\\s*Series\\s*,"))[1]

  if (is.na(header_index)) {
    stop("Could not find a 'Data Series' or 'DataSeries' header in: ", path)
  }

  raw_data <- readr::read_csv(
    path,
    skip = header_index - 1,
    na = c("", "na", "NA", "N.A.", "-"),
    show_col_types = FALSE
  ) |>
    rename_with(trimws)

  names(raw_data)[1] <- "data_series"

  raw_data |>
    mutate(
      row_id = row_number(),
      data_series_raw = data_series,
      data_series = stringr::str_squish(data_series)
    ) |>
    filter(!is.na(data_series), data_series != "")
}

to_long_years <- function(data) {
  year_cols <- names(data)[stringr::str_detect(names(data), "^\\d{4}$")]

  data |>
    pivot_longer(
      cols = all_of(year_cols),
      names_to = "year",
      values_to = "value"
    ) |>
    mutate(
      year = as.integer(year),
      value = readr::parse_number(as.character(value))
    ) |>
    filter(!is.na(year))
}

standardise_age_group <- function(x) {
  x |>
    stringr::str_squish() |>
    stringr::str_replace_all("\\s*-\\s*", "-")
}

is_age_group <- function(x) {
  stringr::str_detect(
    x,
    "^\\d{2}\\s*-\\s*\\d{2}\\s*Years$|^\\d{2}\\s*Years & Over$|^\\d{2}\\s*Years And Over$"
  )
}
```

## 4. Shared age-group definitions

```{r}
asfr_age_groups <- c(
  "15-19 Years",
  "20-24 Years",
  "25-29 Years",
  "30-34 Years",
  "35-39 Years",
  "40-44 Years",
  "45-49 Years"
)

driver_age_groups <- c(
  "25-29 Years",
  "30-34 Years",
  "35-39 Years",
  "40-44 Years",
  "45-49 Years"
)

tibble(
  purpose = c(rep("ASFR", length(asfr_age_groups)), rep("Drivers", length(driver_age_groups))),
  age_group = c(asfr_age_groups, driver_age_groups)
)
```

## 5. Births and fertility rates

```{r}
fertility_wide <- read_singstat_wide(fertility_file)

fertility_wide |>
  select(row_id, data_series) |>
  slice_head(n = 10)
```

```{r}
tfr_annual <- fertility_wide |>
  filter(data_series == "Total Fertility Rate (TFR)") |>
  to_long_years() |>
  transmute(
    year,
    tfr = value
  )

tfr_annual |>
  arrange(year) |>
  slice_head(n = 10)
```

```{r}
asfr_by_age <- fertility_wide |>
  mutate(age_group = standardise_age_group(data_series)) |>
  filter(age_group %in% asfr_age_groups) |>
  to_long_years() |>
  transmute(
    year,
    age_group = factor(age_group, levels = asfr_age_groups),
    asfr = value
  ) |>
  arrange(year, age_group)

asfr_by_age |>
  slice_head(n = 14)
```

## 6. Female singlehood by age group

```{r}
singlehood_wide <- read_singstat_wide(singlehood_file)

singlehood_wide |>
  select(row_id, data_series) |>
  slice_head(n = 20)
```

```{r}
singlehood_tagged <- singlehood_wide |>
  mutate(
    age_group = if_else(
      is_age_group(data_series),
      standardise_age_group(data_series),
      NA_character_
    ),
    sex = case_when(
      data_series == "Males" ~ "Male",
      data_series == "Females" ~ "Female",
      TRUE ~ "Total"
    )
  ) |>
  tidyr::fill(age_group, .direction = "down")

singlehood_tagged |>
  select(row_id, data_series, age_group, sex) |>
  slice_head(n = 20)
```

```{r}
singlehood_female_by_age <- singlehood_tagged |>
  filter(sex == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  transmute(
    year,
    age_group = factor(age_group, levels = driver_age_groups),
    single_female_pct = value
  ) |>
  arrange(year, age_group)

singlehood_female_by_age |>
  slice_head(n = 15)
```

```{r}
singlehood_female_25_49 <- singlehood_female_by_age |>
  group_by(year) |>
  summarise(
    single_female_25_49_pct = mean(single_female_pct, na.rm = TRUE),
    .groups = "drop"
  )

singlehood_female_25_49 |>
  arrange(year) |>
  slice_head(n = 10)
```

## 7. Female labour-force participation rate by age group

```{r}
lfpr_wide <- read_singstat_wide(lfpr_file)

lfpr_wide |>
  select(row_id, data_series) |>
  slice_head(n = 25)
```

```{r}
lfpr_tagged <- lfpr_wide |>
  mutate(
    sex_section = case_when(
      data_series == "Total Resident Participation Rate" ~ "Total",
      data_series == "Male Resident Participation Rate" ~ "Male",
      data_series == "Female Resident Participation Rate" ~ "Female",
      TRUE ~ NA_character_
    ),
    age_group = if_else(
      is_age_group(data_series),
      standardise_age_group(data_series),
      NA_character_
    )
  ) |>
  tidyr::fill(sex_section, .direction = "down")

lfpr_tagged |>
  select(row_id, data_series, sex_section, age_group) |>
  filter(sex_section == "Female") |>
  slice_head(n = 20)
```

```{r}
lfpr_female_by_age <- lfpr_tagged |>
  filter(sex_section == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  transmute(
    year,
    age_group = factor(age_group, levels = driver_age_groups),
    female_lfpr_pct = value
  ) |>
  arrange(year, age_group)

lfpr_female_by_age |>
  slice_head(n = 15)
```

```{r}
lfpr_female_25_49 <- lfpr_female_by_age |>
  group_by(year) |>
  summarise(
    female_lfpr_25_49_pct = mean(female_lfpr_pct, na.rm = TRUE),
    .groups = "drop"
  )

lfpr_female_25_49 |>
  arrange(year) |>
  slice_head(n = 10)
```

## 8. Editable context tables

These policy and recession rows are working annotations for now. We can update them later after the team confirms the final story and references.

```{r}
policy_events <- tribble(
  ~year, ~event_type, ~event_label,
  1972, "policy", "Stop at Two campaign",
  1987, "policy", "Have Three or More if You Can Afford It",
  2001, "policy", "Baby Bonus introduced",
  2013, "policy", "Marriage and Parenthood package enhancement"
)

recession_events <- tribble(
  ~year, ~event_type, ~event_label,
  1985, "recession", "1985 recession",
  1998, "recession", "Asian financial crisis",
  2001, "recession", "2001 downturn",
  2009, "recession", "Global financial crisis",
  2020, "recession", "COVID-19 recession"
)

context_events <- bind_rows(policy_events, recession_events)

context_events
```

## 9. Visualization-ready tables

```{r}
annual_drivers <- tibble(year = 1960:2025) |>
  left_join(tfr_annual, by = "year") |>
  left_join(singlehood_female_25_49, by = "year") |>
  left_join(lfpr_female_25_49, by = "year") |>
  arrange(year)

annual_drivers |>
  slice_head(n = 15)
```

```{r}
asfr_visualization_ready <- asfr_by_age |>
  left_join(annual_drivers, by = "year") |>
  arrange(year, age_group)

asfr_visualization_ready |>
  slice_head(n = 20)
```

## 10. Validation checks

```{r}
validation_summary <- list(
  tfr_year_range = range(tfr_annual$year, na.rm = TRUE),
  asfr_year_range = range(asfr_by_age$year, na.rm = TRUE),
  singlehood_year_range = range(singlehood_female_by_age$year, na.rm = TRUE),
  lfpr_year_range = range(lfpr_female_by_age$year, na.rm = TRUE),
  asfr_age_groups = levels(asfr_by_age$age_group),
  driver_age_groups = driver_age_groups,
  missing_annual_drivers = annual_drivers |>
    summarise(
      missing_tfr = sum(is.na(tfr)),
      missing_singlehood = sum(is.na(single_female_25_49_pct)),
      missing_lfpr = sum(is.na(female_lfpr_25_49_pct))
    )
)

validation_summary
```

```{r}
annual_drivers |>
  summarise(
    min_year = min(year),
    max_year = max(year),
    tfr_available = sum(!is.na(tfr)),
    singlehood_available = sum(!is.na(single_female_25_49_pct)),
    lfpr_available = sum(!is.na(female_lfpr_25_49_pct))
  )
```

## 11. Quick draft chart for checking

This is only a rough check that the cleaned data behaves as expected. We can design the final visualization later.

```{r}
annual_drivers |>
  ggplot(aes(x = year)) +
  geom_line(aes(y = tfr), linewidth = 1, colour = "#1f77b4") +
  labs(
    title = "Draft check: Total Fertility Rate over time",
    x = "Year",
    y = "TFR"
  ) +
  theme_minimal()
```

```{r}
asfr_by_age |>
  ggplot(aes(x = year, y = asfr, fill = age_group)) +
  geom_col() +
  labs(
    title = "Draft check: ASFR by mother's age group",
    x = "Year",
    y = "ASFR",
    fill = "Mother's age group"
  ) +
  theme_minimal()
```

