---
title: "Final Insight Plot Testing"
subtitle: "Singapore fertility visualisation working plots"
author: "Team 11"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## 1. Setup and data engineering

```{r}
library(tidyverse)
```

```{r}
data_dir <- file.path("proposal_files", "datasets")

fertility_file <- file.path(data_dir, "BirthsAndFertilityRatesAnnual.csv")
singlehood_file <- file.path(data_dir, "M810581.csv")
lfpr_file <- file.path(data_dir, "M182201.csv")
family_size_file <- file.path(data_dir, "ResidentEverMarriedFemalesByAgeGroupAndNumberOfChildrenBornAnnual.csv")
first_marriage_file <- file.path(data_dir, "MedianAgeAtFirstMarriageOfResidentCitizenGroomsAndBridesAnnual.csv")
first_birth_file <- file.path(data_dir, "MedianAgeOfMothersAtFirstBirthAnnual.csv")
cpi_file <- file.path(data_dir, "ConsumerPriceIndexCPI2024AsBaseYearMonthly.csv")
education_file <- file.path(data_dir, "M850581.csv")
occupation_file <- file.path(data_dir, "EmployedResidentsAged15YearsAndOverByOccupationAndAgeGroupJuneAnnual.csv")

tibble(
  dataset = c(
    "fertility",
    "singlehood",
    "female_lfpr",
    "family_size",
    "first_marriage",
    "first_birth",
    "cpi",
    "education",
    "occupation"
  ),
  file = c(
    fertility_file,
    singlehood_file,
    lfpr_file,
    family_size_file,
    first_marriage_file,
    first_birth_file,
    cpi_file,
    education_file,
    occupation_file
  ),
  exists = file.exists(c(
    fertility_file,
    singlehood_file,
    lfpr_file,
    family_size_file,
    first_marriage_file,
    first_birth_file,
    cpi_file,
    education_file,
    occupation_file
  ))
)
```

```{r}
survey_barriers_2021 <- tribble(
  ~reason, ~share_pct,
  "Financial cost", 64,
  "Already have enough children", 50,
  "Raising children in Singapore is too stressful", 49,
  "Difficult to manage work and family demands", 34,
  "Lack of good caregiving arrangements", 17,
  "Difficulty conceiving", 14
) |>
  mutate(
    reason = factor(reason, levels = rev(reason)),
    barrier_type = if_else(
      reason %in% c("Financial cost", "Raising children in Singapore is too stressful"),
      "Cost or stress",
      "Other reason"
    )
  )
```

```{r}
cpi_monthly <- readr::read_csv(
  cpi_file,
  na = c("", "na", "NA", "N.A.", "-"),
  show_col_types = FALSE
) |>
  rename(data_series = DataSeries) |>
  mutate(data_series = stringr::str_squish(data_series)) |>
  filter(data_series == "All Items") |>
  pivot_longer(
    cols = -data_series,
    names_to = "month_code",
    values_to = "cpi_index"
  ) |>
  mutate(
    year = as.integer(stringr::str_sub(month_code, 1, 4)),
    month = match(stringr::str_sub(month_code, 5, 7), month.abb),
    cpi_index = readr::parse_number(as.character(cpi_index))
  ) |>
  filter(!is.na(year), !is.na(month), !is.na(cpi_index))

cpi_annual <- cpi_monthly |>
  filter(year >= 1990, year <= 2025) |>
  group_by(year) |>
  summarise(cpi_index = mean(cpi_index, na.rm = TRUE), .groups = "drop") |>
  mutate(
    cpi_1990_100 = cpi_index / cpi_index[year == 1990] * 100
  )
```

```{r}
read_singstat_wide <- function(path) {
  lines <- readLines(path, warn = FALSE)
  header_index <- which(stringr::str_detect(lines, "^Data\\s*Series\\s*,"))[1]

  if (is.na(header_index)) {
    stop("Could not find a Data Series header in: ", path)
  }

  raw_data <- readr::read_csv(
    path,
    skip = header_index - 1,
    na = c("", "na", "NA", "N.A.", "-"),
    show_col_types = FALSE
  ) |>
    rename_with(trimws)

  names(raw_data)[1] <- "data_series"

  raw_data |>
    mutate(
      row_id = row_number(),
      data_series_raw = data_series,
      data_series = stringr::str_squish(data_series)
    ) |>
    filter(!is.na(data_series), data_series != "")
}

to_long_years <- function(data) {
  year_cols <- names(data)[stringr::str_detect(names(data), "^\\d{4}$")]

  data |>
    pivot_longer(
      cols = all_of(year_cols),
      names_to = "year",
      values_to = "value"
    ) |>
    mutate(
      year = as.integer(year),
      value = readr::parse_number(as.character(value))
    ) |>
    filter(!is.na(year))
}

standardise_age_group <- function(x) {
  x |>
    stringr::str_squish() |>
    stringr::str_replace_all("\\s*-\\s*", "-")
}

is_age_group <- function(x) {
  stringr::str_detect(
    x,
    "^\\d{2}\\s*-\\s*\\d{2}\\s*Years$|^\\d{2}\\s*Years & Over$|^\\d{2}\\s*Years And Over$"
  )
}
```

```{r}
asfr_age_groups <- c(
  "15-19 Years", "20-24 Years", "25-29 Years", "30-34 Years",
  "35-39 Years", "40-44 Years", "45-49 Years"
)

driver_age_groups <- c(
  "25-29 Years", "30-34 Years", "35-39 Years", "40-44 Years", "45-49 Years"
)

children_born_levels <- c(
  "None", "1 Child", "2 Children", "3 Children", "4 Or More Children"
)
```

```{r}
fertility_wide <- read_singstat_wide(fertility_file)

tfr_annual <- fertility_wide |>
  filter(data_series == "Total Fertility Rate (TFR)") |>
  to_long_years() |>
  transmute(year, tfr = value)

asfr_by_age <- fertility_wide |>
  mutate(age_group = standardise_age_group(data_series)) |>
  filter(age_group %in% asfr_age_groups) |>
  to_long_years() |>
  transmute(
    year,
    age_group = factor(age_group, levels = asfr_age_groups),
    asfr = value
  )
```

```{r}
singlehood_wide <- read_singstat_wide(singlehood_file)

singlehood_tagged <- singlehood_wide |>
  mutate(
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_),
    sex = case_when(
      data_series == "Males" ~ "Male",
      data_series == "Females" ~ "Female",
      TRUE ~ "Total"
    )
  ) |>
  tidyr::fill(age_group, .direction = "down")

singlehood_female_25_49 <- singlehood_tagged |>
  filter(sex == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  group_by(year) |>
  summarise(single_female_25_49_pct = mean(value, na.rm = TRUE), .groups = "drop")
```

```{r}
lfpr_wide <- read_singstat_wide(lfpr_file)

lfpr_tagged <- lfpr_wide |>
  mutate(
    sex_section = case_when(
      data_series == "Total Resident Participation Rate" ~ "Total",
      data_series == "Male Resident Participation Rate" ~ "Male",
      data_series == "Female Resident Participation Rate" ~ "Female",
      TRUE ~ NA_character_
    ),
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_)
  ) |>
  tidyr::fill(sex_section, .direction = "down")

lfpr_female_25_49 <- lfpr_tagged |>
  filter(sex_section == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_lfpr_25_49_pct = mean(value, na.rm = TRUE), .groups = "drop")
```

```{r}
education_wide <- read_singstat_wide(education_file)

female_total_25_49 <- education_wide |>
  filter(
    row_id > row_id[data_series == "Total Females"][1],
    row_id < row_id[data_series == "Total Females - Below Secondary"][1],
    standardise_age_group(data_series) %in% driver_age_groups
  ) |>
  mutate(age_group = standardise_age_group(data_series)) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_total_25_49 = sum(value, na.rm = TRUE), .groups = "drop")

female_university_25_49 <- education_wide |>
  filter(
    row_id > row_id[data_series == "Total Females - University"][1],
    standardise_age_group(data_series) %in% driver_age_groups
  ) |>
  mutate(age_group = standardise_age_group(data_series)) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_university_25_49 = sum(value, na.rm = TRUE), .groups = "drop")

female_university_share_25_49 <- female_total_25_49 |>
  left_join(female_university_25_49, by = "year") |>
  mutate(
    female_university_25_49_pct = female_university_25_49 / female_total_25_49 * 100
  )
```

```{r}
occupation_wide <- read_singstat_wide(occupation_file)

pmet_series <- c(
  "Managers & Administrators (Including Working Proprietors)",
  "Professionals",
  "Associate Professionals & Technicians"
)

pmet_share <- occupation_wide |>
  filter(
    row_id < row_id[data_series == "Employed Residents Aged 15 - 19 Years"][1],
    data_series %in% c("All Occupation Groups, (Total Employed Residents)", pmet_series)
  ) |>
  to_long_years() |>
  mutate(
    occupation_group = case_when(
      data_series == "All Occupation Groups, (Total Employed Residents)" ~ "Total employed residents",
      data_series %in% pmet_series ~ "PMET",
      TRUE ~ NA_character_
    )
  ) |>
  group_by(year, occupation_group) |>
  summarise(employed_residents = sum(value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = occupation_group, values_from = employed_residents) |>
  mutate(
    pmet_share_pct = PMET / `Total employed residents` * 100
  ) |>
  filter(year >= 2000, year <= 2025)
```

```{r}
annual_drivers <- tibble(year = 1960:2025) |>
  left_join(tfr_annual, by = "year") |>
  left_join(singlehood_female_25_49, by = "year") |>
  left_join(lfpr_female_25_49, by = "year") |>
  mutate(decade = paste0(floor(year / 10) * 10, "s"))
```

```{r}
family_size_wide <- read_singstat_wide(family_size_file)

family_size_tagged <- family_size_wide |>
  mutate(
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_),
    children_born = if_else(data_series %in% children_born_levels, data_series, NA_character_)
  ) |>
  tidyr::fill(age_group, .direction = "down")

family_size_40_49 <- family_size_tagged |>
  filter(age_group == "40-49 Years", children_born %in% children_born_levels) |>
  to_long_years() |>
  transmute(
    year,
    children_born = factor(children_born, levels = children_born_levels),
    females = value
  ) |>
  group_by(year) |>
  mutate(
    total_females = sum(females, na.rm = TRUE),
    share_pct = females / total_females * 100
  ) |>
  ungroup()
```

```{r}
first_marriage_wide <- read_singstat_wide(first_marriage_file)

resident_brides_first_marriage <- first_marriage_wide |>
  filter(data_series == "Resident Brides") |>
  to_long_years() |>
  transmute(
    year,
    median_age_first_marriage = value
  )
```

```{r}
first_birth_wide <- read_singstat_wide(first_birth_file)

resident_first_birth_age <- first_birth_wide |>
  filter(data_series == "Median Age Of Mothers At First Birth For Resident Births") |>
  to_long_years() |>
  transmute(
    year,
    median_age_first_birth = value
  )
```

## 2. Insight 1 plot: did policies and recessions explain sustained fertility changes?

```{r}
policy_events <- tribble(
  ~event_year, ~event_label, ~policy_stance,
  1972, "Stop at Two", "Anti-natalist",
  1987, "Have Three or More", "Pro-natalist",
  2001, "Baby Bonus", "Pro-natalist",
  2013, "Marriage and Parenthood package", "Pro-natalist"
)

economic_shocks <- tribble(
  ~shock_year, ~shock_label, ~severity, ~quarters_below_peak,
  1985, "1985", "Severe", 6,
  1998, "1998", "Severe", 6,
  2001, "2001", "Moderate", 5,
  2009, "2009", "Most severe", 6,
  2020, "2020", "Most severe", 4
)

policy_window <- policy_events |>
  crossing(relative_year = -5:10) |>
  mutate(year = event_year + relative_year) |>
  left_join(tfr_annual, by = "year") |>
  mutate(
    event_label = factor(
      event_label,
      levels = c("Stop at Two", "Have Three or More", "Baby Bonus", "Marriage and Parenthood package")
    )
  )

shock_window <- policy_events |>
  crossing(economic_shocks) |>
  mutate(
    relative_year = shock_year - event_year,
    recovery_years = quarters_below_peak / 4,
    relative_recovery_end = relative_year + recovery_years,
    event_label = factor(
      event_label,
      levels = c("Stop at Two", "Have Three or More", "Baby Bonus", "Marriage and Parenthood package")
    )
  ) |>
  filter(relative_year <= 10, relative_recovery_end >= -5) |>
  mutate(
    xmin = pmax(relative_year, -5),
    xmax = pmin(relative_recovery_end, 10)
  )

ggplot(policy_window, aes(x = relative_year, y = tfr, colour = policy_stance)) +
  geom_rect(
    data = shock_window,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = -Inf,
      ymax = Inf,
      fill = severity
    ),
    inherit.aes = FALSE,
    alpha = 0.24
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey35") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.3) +
  geom_label(
    data = shock_window,
    aes(x = (xmin + xmax) / 2, y = Inf, label = shock_label),
    inherit.aes = FALSE,
    vjust = 1.2,
    size = 2.8,
    label.size = 0.15,
    fill = "white",
    colour = "grey35"
  ) +
  facet_wrap(~ event_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("Anti-natalist" = "#D95F02", "Pro-natalist" = "#1B9E77")) +
  scale_fill_manual(
    values = c(
      "Most severe" = "#B2182B",
      "Severe" = "#8073AC",
      "Moderate" = "#D8DAEB"
    )
  ) +
  scale_x_continuous(breaks = c(-5, 0, 5, 10)) +
  labs(
    title = "Did policies and recessions explain sustained fertility changes?",
    subtitle = "Dashed line = policy year. Shaded bands show economic recovery duration, with warmer colours indicating higher severity.",
    x = "Years from policy",
    y = "TFR",
    colour = "Policy stance",
    fill = "Economic shock severity",
    caption = "Band width is based on quarters below the previous GDP peak. Timing is context only, not causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```

## 3. Insight 2 plot: did female LFPR rise with demand for professional work?

```{r}
pmet_lfpr <- pmet_share |>
  left_join(lfpr_female_25_49, by = "year") |>
  filter(
    year >= 2000,
    year <= 2025,
    !is.na(pmet_share_pct),
    !is.na(female_lfpr_25_49_pct)
  ) |>
  mutate(decade = paste0(floor(year / 10) * 10, "s"))

pmet_lfpr_labels <- pmet_lfpr |>
  filter(year %in% c(2000, 2010, 2020, 2025))

ggplot(
  pmet_lfpr,
  aes(
    x = pmet_share_pct,
    y = female_lfpr_25_49_pct,
    colour = year
  )
) +
  geom_path(aes(group = 1), colour = "grey65", linewidth = 0.9, alpha = 0.85) +
  geom_point(size = 4.6, alpha = 0.95) +
  geom_label(
    data = pmet_lfpr_labels,
    aes(label = year),
    nudge_y = 1.0,
    size = 3.5,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(40, 70),
    breaks = seq(40, 70, 5)
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(65, 90),
    breaks = seq(65, 90, 5)
  ) +
  scale_colour_gradient(
    low = "#9ECAE1",
    high = "#08519C",
    breaks = c(2000, 2010, 2020, 2025)
  ) +
  labs(
    title = "Did female LFPR rise with demand for professional work?",
    subtitle = "Moving right = higher PMET share. Moving up = higher female LFPR among women aged 25 to 49.",
    x = "PMET share of employed residents",
    y = "Resident female LFPR, ages 25 to 49",
    colour = "Year",
    caption = "PMET refers to managers, professionals, associate professionals and technicians. This is descriptive context, not causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 4. Insight 3 plot A: did higher female education coincide with fewer births in the early twenties?

```{r}
asfr_20_24 <- asfr_by_age |>
  filter(age_group == "20-24 Years") |>
  transmute(year, asfr_20_24 = asfr)

education_timing <- female_university_share_25_49 |>
  left_join(asfr_20_24, by = "year") |>
  left_join(resident_brides_first_marriage, by = "year") |>
  left_join(resident_first_birth_age, by = "year") |>
  filter(
    year >= 2000,
    year <= 2025,
    !is.na(female_university_25_49_pct)
  ) |>
  mutate(decade = paste0(floor(year / 10) * 10, "s"))

education_asfr <- education_timing |>
  filter(!is.na(asfr_20_24))

education_asfr_labels <- education_asfr |>
  filter(year %in% c(2000, 2010, 2020, 2025))

ggplot(
  education_asfr,
  aes(
    x = female_university_25_49_pct,
    y = asfr_20_24,
    colour = year
  )
) +
  geom_path(aes(group = 1), colour = "grey65", linewidth = 0.9, alpha = 0.85) +
  geom_point(size = 4.6, alpha = 0.95) +
  geom_label(
    data = education_asfr_labels,
    aes(label = year),
    nudge_y = 8,
    size = 3.5,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    breaks = seq(15, 60, 5)
  ) +
  scale_y_continuous(
    breaks = seq(0, 80, 10),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  scale_colour_gradient(
    low = "#FDD49E",
    high = "#D95F0E",
    breaks = c(2000, 2010, 2020, 2025)
  ) +
  labs(
    title = "Did higher female education coincide with fewer births in the early twenties?",
    subtitle = "Moving right = higher female university share. Moving down = lower ASFR among women aged 20 to 24.",
    x = "Female university share, ages 25 to 49",
    y = "ASFR, women aged 20 to 24",
    colour = "Year",
    caption = "ASFR means age-specific fertility rate. This supports the timing-shift story, not direct causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 5. Insight 3 plot B: did demand for professional work rise alongside female education?

```{r}
education_pmet <- pmet_share |>
  left_join(female_university_share_25_49, by = "year") |>
  filter(
    year >= 2000,
    year <= 2025,
    !is.na(pmet_share_pct),
    !is.na(female_university_25_49_pct)
  )

education_pmet_labels <- education_pmet |>
  filter(year %in% c(2000, 2010, 2020, 2025))

ggplot(
  education_pmet,
  aes(
    x = pmet_share_pct,
    y = female_university_25_49_pct,
    colour = year
  )
) +
  geom_path(aes(group = 1), colour = "grey65", linewidth = 0.9, alpha = 0.85) +
  geom_point(size = 4.6, alpha = 0.95) +
  geom_label(
    data = education_pmet_labels,
    aes(label = year),
    nudge_y = 2.0,
    size = 3.5,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(40, 70),
    breaks = seq(40, 70, 5)
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    breaks = seq(15, 60, 5),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  scale_colour_gradient(
    low = "#BFD3E6",
    high = "#08519C",
    breaks = c(2000, 2010, 2020, 2025)
  ) +
  labs(
    title = "Did demand for professional work rise alongside female education?",
    subtitle = "Moving right = higher PMET share. Moving up = higher female university share among women aged 25 to 49.",
    x = "PMET share of employed residents",
    y = "Female university share, ages 25 to 49",
    colour = "Year",
    caption = "PMET share is used as a proxy for a more professional labour market. This is descriptive context, not causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 6. Insight 3 plot C: did higher female education coincide with later marriage and first birth?

```{r}
education_timing_long <- bind_rows(
  education_timing |>
    transmute(
      year,
      female_university_25_49_pct,
      measure = "First marriage age",
      median_age = median_age_first_marriage
    ),
  education_timing |>
    transmute(
      year,
      female_university_25_49_pct,
      measure = "First birth age",
      median_age = median_age_first_birth
    )
) |>
  filter(!is.na(median_age), !is.na(female_university_25_49_pct)) |>
  mutate(
    measure = factor(measure, levels = c("First marriage age", "First birth age"))
  )

education_timing_labels <- education_timing_long |>
  group_by(measure) |>
  filter(year %in% c(min(year), max(year))) |>
  ungroup() |>
  mutate(
    label = paste0(year, ": ", round(median_age, 1)),
    label_y = if_else(measure == "First birth age", median_age + 0.18, median_age - 0.18),
    hjust_label = if_else(year == min(year), 1.08, -0.08)
  )

ggplot(
  education_timing_long,
  aes(
    x = female_university_25_49_pct,
    y = median_age,
    colour = measure
  )
) +
  geom_path(aes(group = measure), linewidth = 1.15, alpha = 0.9) +
  geom_point(size = 3.4, alpha = 0.95) +
  geom_label(
    data = education_timing_labels,
    aes(y = label_y, label = label, hjust = hjust_label),
    size = 3.3,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    breaks = seq(15, 60, 5),
    expand = expansion(mult = c(0.18, 0.18))
  ) +
  scale_y_continuous(
    breaks = seq(25, 33, 1),
    expand = expansion(mult = c(0.08, 0.12))
  ) +
  scale_colour_manual(
    values = c("First marriage age" = "#2F4B7C", "First birth age" = "#7A5195")
  ) +
  labs(
    title = "Did higher female education coincide with later marriage and first birth?",
    subtitle = "Moving right = higher female university share. Moving up = later family formation.",
    x = "Female university share, ages 25 to 49",
    y = "Median age",
    colour = NULL,
    caption = "This describes timing patterns. It does not prove education directly caused later marriage or birth."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 7. Insight 4 plot A: did family formation shift later while large families became less common?

```{r}
large_family_share <- family_size_40_49 |>
  mutate(
    family_size_group = case_when(
      children_born %in% c("3 Children", "4 Or More Children") ~ "3 or more children",
      TRUE ~ "Other"
    )
  ) |>
  filter(family_size_group == "3 or more children") |>
  group_by(year) |>
  summarise(large_family_share_pct = sum(share_pct, na.rm = TRUE), .groups = "drop")

family_formation_timeline <- bind_rows(
  resident_brides_first_marriage |>
    filter(year >= 1990, year <= 2025) |>
    transmute(
      year,
      indicator = "First marriage age",
      value = median_age_first_marriage,
      unit = "years"
    ),
  resident_first_birth_age |>
    filter(year >= 1990, year <= 2025) |>
    transmute(
      year,
      indicator = "First birth age",
      value = median_age_first_birth,
      unit = "years"
    ),
  singlehood_female_25_49 |>
    filter(year >= 1990, year <= 2025) |>
    transmute(
      year,
      indicator = "Single women aged 25 to 49",
      value = single_female_25_49_pct,
      unit = "%"
    ),
  large_family_share |>
    filter(year >= 1990, year <= 2025) |>
    transmute(
      year,
      indicator = "Ever-married women aged 40 to 49 with 3+ children",
      value = large_family_share_pct,
      unit = "%"
    )
) |>
  mutate(
    indicator = factor(
      indicator,
      levels = c(
        "First marriage age",
        "First birth age",
        "Single women aged 25 to 49",
        "Ever-married women aged 40 to 49 with 3+ children"
      )
    ),
    direction = if_else(
      indicator == "Ever-married women aged 40 to 49 with 3+ children",
      "Larger families became less common",
      "Family formation shifted later"
    )
  )

timeline_labels <- family_formation_timeline |>
  group_by(indicator) |>
  filter(year == max(year)) |>
  ungroup() |>
  mutate(
    label = if_else(
      unit == "%",
      paste0(round(value, 1), "%"),
      paste0(round(value, 1), " yrs")
    ),
    label = paste0(year, ": ", label),
    label_y = case_when(
      indicator == "First marriage age" ~ value + 0.28,
      indicator == "First birth age" ~ value - 0.28,
      indicator == "Single women aged 25 to 49" ~ value + 0.85,
      indicator == "Ever-married women aged 40 to 49 with 3+ children" ~ value + 1.25,
      TRUE ~ value
    )
  )

ggplot(family_formation_timeline, aes(x = year, y = value, colour = direction)) +
  geom_line(linewidth = 1.15) +
  geom_point(
    data = timeline_labels,
    fill = "white",
    shape = 21,
    size = 3.3,
    stroke = 1
  ) +
  geom_text(
    data = timeline_labels,
    aes(y = label_y, label = label),
    hjust = 0,
    nudge_x = 0.35,
    size = 3.3,
    fontface = "bold",
    colour = "grey25",
    show.legend = FALSE
  ) +
  facet_wrap(~ indicator, scales = "free_y", ncol = 1) +
  scale_colour_manual(
    values = c(
      "Family formation shifted later" = "#2F4B7C",
      "Larger families became less common" = "#C94B55"
    )
  ) +
  scale_x_continuous(
    breaks = seq(1990, 2025, 5),
    expand = expansion(mult = c(0.04, 0.18))
  ) +
  labs(
    title = "Did family formation shift later while large families became less common?",
    subtitle = "All panels use the same time axis. Values use separate y-scales because the units differ.",
    x = "Year",
    y = NULL,
    colour = NULL,
    caption = "The timeline shows connected descriptive patterns, not direct causality."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", hjust = 0),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 8. Insight 4 plot B: which family sizes became more or less common?

```{r}
family_change <- family_size_40_49 |>
  mutate(
    family_size_group = case_when(
      children_born == "None" ~ "None",
      children_born == "1 Child" ~ "1 child",
      children_born == "2 Children" ~ "2 children",
      children_born %in% c("3 Children", "4 Or More Children") ~ "3 or more children",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(family_size_group), year %in% c(1980, 2025)) |>
  group_by(year, family_size_group) |>
  summarise(share_pct = sum(share_pct, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = share_pct, names_prefix = "share_") |>
  mutate(
    change_pp = share_2025 - share_1980,
    direction = if_else(change_pp >= 0, "Became more common", "Became less common"),
    label = paste0(if_else(change_pp >= 0, "+", ""), round(change_pp, 1), " pp"),
    family_size_group = factor(
      family_size_group,
      levels = c("None", "1 child", "2 children", "3 or more children")
    )
  )

ggplot(family_change, aes(x = change_pp, y = family_size_group, fill = direction)) +
  geom_vline(xintercept = 0, colour = "grey45") +
  geom_col(width = 0.66) +
  geom_text(
    aes(
      label = label,
      hjust = if_else(change_pp >= 0, -0.12, 1.12)
    ),
    size = 3.8,
    fontface = "bold",
    colour = "grey20"
  ) +
  scale_x_continuous(
    labels = function(x) paste0(x, " pp"),
    expand = expansion(mult = c(0.18, 0.18))
  ) +
  scale_fill_manual(values = c("Became more common" = "#1B9E77", "Became less common" = "#C94B55")) +
  labs(
    title = "Which family sizes became more or less common?",
    subtitle = "Change in share of ever-married resident females aged 40 to 49, 1980 to 2025.",
    x = "Change in share, percentage points",
    y = "Number of children born",
    fill = NULL,
    caption = "This measures near-completed family size."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )
```
