---
title: "Final Insight Plot Testing"
subtitle: "Singapore fertility visualisation working plots"
author: "Team 11"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## 1. Setup and data engineering

```{r}
library(tidyverse)
```

```{r}
data_dir <- file.path("proposal_files", "datasets")

fertility_file <- file.path(data_dir, "BirthsAndFertilityRatesAnnual.csv")
singlehood_file <- file.path(data_dir, "M810581.csv")
lfpr_file <- file.path(data_dir, "M182201.csv")
family_size_file <- file.path(data_dir, "ResidentEverMarriedFemalesByAgeGroupAndNumberOfChildrenBornAnnual.csv")
first_marriage_file <- file.path(data_dir, "MedianAgeAtFirstMarriageOfResidentCitizenGroomsAndBridesAnnual.csv")
first_birth_file <- file.path(data_dir, "MedianAgeOfMothersAtFirstBirthAnnual.csv")
cpi_file <- file.path(data_dir, "ConsumerPriceIndexCPI2024AsBaseYearMonthly.csv")
education_file <- file.path(data_dir, "M850581.csv")
occupation_file <- file.path(data_dir, "EmployedResidentsAged15YearsAndOverByOccupationAndAgeGroupJuneAnnual.csv")

tibble(
  dataset = c(
    "fertility",
    "singlehood",
    "female_lfpr",
    "family_size",
    "first_marriage",
    "first_birth",
    "cpi",
    "education",
    "occupation"
  ),
  file = c(
    fertility_file,
    singlehood_file,
    lfpr_file,
    family_size_file,
    first_marriage_file,
    first_birth_file,
    cpi_file,
    education_file,
    occupation_file
  ),
  exists = file.exists(c(
    fertility_file,
    singlehood_file,
    lfpr_file,
    family_size_file,
    first_marriage_file,
    first_birth_file,
    cpi_file,
    education_file,
    occupation_file
  ))
)
```

```{r}
survey_barriers_2021 <- tribble(
  ~reason, ~share_pct,
  "Financial cost", 64,
  "Already have enough children", 50,
  "Raising children in Singapore is too stressful", 49,
  "Difficult to manage work and family demands", 34,
  "Lack of good caregiving arrangements", 17,
  "Difficulty conceiving", 14
) |>
  mutate(
    reason = factor(reason, levels = rev(reason)),
    barrier_type = if_else(
      reason %in% c("Financial cost", "Raising children in Singapore is too stressful"),
      "Cost or stress",
      "Other reason"
    )
  )
```

```{r}
cpi_monthly <- readr::read_csv(
  cpi_file,
  na = c("", "na", "NA", "N.A.", "-"),
  show_col_types = FALSE
) |>
  rename(data_series = DataSeries) |>
  mutate(data_series = stringr::str_squish(data_series)) |>
  filter(data_series == "All Items") |>
  pivot_longer(
    cols = -data_series,
    names_to = "month_code",
    values_to = "cpi_index"
  ) |>
  mutate(
    year = as.integer(stringr::str_sub(month_code, 1, 4)),
    month = match(stringr::str_sub(month_code, 5, 7), month.abb),
    cpi_index = readr::parse_number(as.character(cpi_index))
  ) |>
  filter(!is.na(year), !is.na(month), !is.na(cpi_index))

cpi_annual <- cpi_monthly |>
  filter(year >= 1990, year <= 2025) |>
  group_by(year) |>
  summarise(cpi_index = mean(cpi_index, na.rm = TRUE), .groups = "drop") |>
  mutate(
    cpi_1990_100 = cpi_index / cpi_index[year == 1990] * 100
  )
```

```{r}
read_singstat_wide <- function(path) {
  lines <- readLines(path, warn = FALSE)
  header_index <- which(stringr::str_detect(lines, "^Data\\s*Series\\s*,"))[1]

  if (is.na(header_index)) {
    stop("Could not find a Data Series header in: ", path)
  }

  raw_data <- readr::read_csv(
    path,
    skip = header_index - 1,
    na = c("", "na", "NA", "N.A.", "-"),
    show_col_types = FALSE
  ) |>
    rename_with(trimws)

  names(raw_data)[1] <- "data_series"

  raw_data |>
    mutate(
      row_id = row_number(),
      data_series_raw = data_series,
      data_series = stringr::str_squish(data_series)
    ) |>
    filter(!is.na(data_series), data_series != "")
}

to_long_years <- function(data) {
  year_cols <- names(data)[stringr::str_detect(names(data), "^\\d{4}$")]

  data |>
    pivot_longer(
      cols = all_of(year_cols),
      names_to = "year",
      values_to = "value"
    ) |>
    mutate(
      year = as.integer(year),
      value = readr::parse_number(as.character(value))
    ) |>
    filter(!is.na(year))
}

standardise_age_group <- function(x) {
  x |>
    stringr::str_squish() |>
    stringr::str_replace_all("\\s*-\\s*", "-")
}

is_age_group <- function(x) {
  stringr::str_detect(
    x,
    "^\\d{2}\\s*-\\s*\\d{2}\\s*Years$|^\\d{2}\\s*Years & Over$|^\\d{2}\\s*Years And Over$"
  )
}
```

```{r}
asfr_age_groups <- c(
  "15-19 Years", "20-24 Years", "25-29 Years", "30-34 Years",
  "35-39 Years", "40-44 Years", "45-49 Years"
)

driver_age_groups <- c(
  "25-29 Years", "30-34 Years", "35-39 Years", "40-44 Years", "45-49 Years"
)

children_born_levels <- c(
  "None", "1 Child", "2 Children", "3 Children", "4 Or More Children"
)
```

```{r}
fertility_wide <- read_singstat_wide(fertility_file)

tfr_annual <- fertility_wide |>
  filter(data_series == "Total Fertility Rate (TFR)") |>
  to_long_years() |>
  transmute(year, tfr = value)

asfr_by_age <- fertility_wide |>
  mutate(age_group = standardise_age_group(data_series)) |>
  filter(age_group %in% asfr_age_groups) |>
  to_long_years() |>
  transmute(
    year,
    age_group = factor(age_group, levels = asfr_age_groups),
    asfr = value
  )
```

```{r}
singlehood_wide <- read_singstat_wide(singlehood_file)

singlehood_tagged <- singlehood_wide |>
  mutate(
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_),
    sex = case_when(
      data_series == "Males" ~ "Male",
      data_series == "Females" ~ "Female",
      TRUE ~ "Total"
    )
  ) |>
  tidyr::fill(age_group, .direction = "down")

singlehood_female_25_49 <- singlehood_tagged |>
  filter(sex == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  group_by(year) |>
  summarise(single_female_25_49_pct = mean(value, na.rm = TRUE), .groups = "drop")
```

```{r}
lfpr_wide <- read_singstat_wide(lfpr_file)

lfpr_tagged <- lfpr_wide |>
  mutate(
    sex_section = case_when(
      data_series == "Total Resident Participation Rate" ~ "Total",
      data_series == "Male Resident Participation Rate" ~ "Male",
      data_series == "Female Resident Participation Rate" ~ "Female",
      TRUE ~ NA_character_
    ),
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_)
  ) |>
  tidyr::fill(sex_section, .direction = "down")

lfpr_female_25_49 <- lfpr_tagged |>
  filter(sex_section == "Female", age_group %in% driver_age_groups) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_lfpr_25_49_pct = mean(value, na.rm = TRUE), .groups = "drop")
```

```{r}
education_wide <- read_singstat_wide(education_file)

female_total_25_49 <- education_wide |>
  filter(
    row_id > row_id[data_series == "Total Females"][1],
    row_id < row_id[data_series == "Total Females - Below Secondary"][1],
    standardise_age_group(data_series) %in% driver_age_groups
  ) |>
  mutate(age_group = standardise_age_group(data_series)) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_total_25_49 = sum(value, na.rm = TRUE), .groups = "drop")

female_university_25_49 <- education_wide |>
  filter(
    row_id > row_id[data_series == "Total Females - University"][1],
    standardise_age_group(data_series) %in% driver_age_groups
  ) |>
  mutate(age_group = standardise_age_group(data_series)) |>
  to_long_years() |>
  group_by(year) |>
  summarise(female_university_25_49 = sum(value, na.rm = TRUE), .groups = "drop")

female_university_share_25_49 <- female_total_25_49 |>
  left_join(female_university_25_49, by = "year") |>
  mutate(
    female_university_25_49_pct = female_university_25_49 / female_total_25_49 * 100
  )
```

```{r}
occupation_wide <- read_singstat_wide(occupation_file)

pmet_series <- c(
  "Managers & Administrators (Including Working Proprietors)",
  "Professionals",
  "Associate Professionals & Technicians"
)

pmet_share <- occupation_wide |>
  filter(
    row_id < row_id[data_series == "Employed Residents Aged 15 - 19 Years"][1],
    data_series %in% c("All Occupation Groups, (Total Employed Residents)", pmet_series)
  ) |>
  to_long_years() |>
  mutate(
    occupation_group = case_when(
      data_series == "All Occupation Groups, (Total Employed Residents)" ~ "Total employed residents",
      data_series %in% pmet_series ~ "PMET",
      TRUE ~ NA_character_
    )
  ) |>
  group_by(year, occupation_group) |>
  summarise(employed_residents = sum(value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = occupation_group, values_from = employed_residents) |>
  mutate(
    pmet_share_pct = PMET / `Total employed residents` * 100
  ) |>
  filter(year >= 2000, year <= 2025)
```

```{r}
annual_drivers <- tibble(year = 1960:2025) |>
  left_join(tfr_annual, by = "year") |>
  left_join(singlehood_female_25_49, by = "year") |>
  left_join(lfpr_female_25_49, by = "year") |>
  mutate(decade = paste0(floor(year / 10) * 10, "s"))
```

```{r}
family_size_wide <- read_singstat_wide(family_size_file)

family_size_tagged <- family_size_wide |>
  mutate(
    age_group = if_else(is_age_group(data_series), standardise_age_group(data_series), NA_character_),
    children_born = if_else(data_series %in% children_born_levels, data_series, NA_character_)
  ) |>
  tidyr::fill(age_group, .direction = "down")

family_size_40_49 <- family_size_tagged |>
  filter(age_group == "40-49 Years", children_born %in% children_born_levels) |>
  to_long_years() |>
  transmute(
    year,
    children_born = factor(children_born, levels = children_born_levels),
    females = value
  ) |>
  group_by(year) |>
  mutate(
    total_females = sum(females, na.rm = TRUE),
    share_pct = females / total_females * 100
  ) |>
  ungroup()
```

```{r}
first_marriage_wide <- read_singstat_wide(first_marriage_file)

resident_brides_first_marriage <- first_marriage_wide |>
  filter(data_series == "Resident Brides") |>
  to_long_years() |>
  transmute(
    year,
    median_age_first_marriage = value
  )
```

```{r}
first_birth_wide <- read_singstat_wide(first_birth_file)

resident_first_birth_age <- first_birth_wide |>
  filter(data_series == "Median Age Of Mothers At First Birth For Resident Births") |>
  to_long_years() |>
  transmute(
    year,
    median_age_first_birth = value
  )
```

## 2. Insight 1 plot A: what reasons did couples give for not having more children?

```{r}
ggplot(survey_barriers_2021, aes(x = share_pct, y = reason, fill = barrier_type)) +
  geom_col(width = 0.64) +
  geom_text(
    aes(label = paste0(share_pct, "%")),
    hjust = -0.12,
    size = 3.8,
    fontface = "bold",
    colour = "grey20"
  ) +
  scale_x_continuous(
    labels = scales::label_percent(scale = 1),
    limits = c(0, 72),
    breaks = seq(0, 70, 10),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c("Cost or stress" = "#C94B55", "Other reason" = "#5A8BB8")
  ) +
  labs(
    title = "What reasons did couples give for not having more children?",
    subtitle = "Married respondents who do not want children or more children, 2021.",
    x = "Selected reason among top three",
    y = NULL,
    fill = NULL,
    caption = "Source: Marriage and Parenthood Survey 2021, Annex B, Table 3. Respondents could choose up to three reasons."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.major.y = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 3. Insight 1 plot B: did cost, female LFPR and singlehood rise together?

```{r}
insight1_path <- cpi_annual |>
  select(year, cpi_1990_100) |>
  left_join(lfpr_female_25_49, by = "year") |>
  left_join(singlehood_female_25_49, by = "year") |>
  filter(
    year >= 1990,
    year <= 2025,
    !is.na(cpi_1990_100),
    !is.na(female_lfpr_25_49_pct),
    !is.na(single_female_25_49_pct)
  )

insight1_path_labels <- insight1_path |>
  filter(year %in% c(1990, 2000, 2010, 2020, 2025)) |>
  mutate(
    label_x = case_when(
      year == 2000 ~ cpi_1990_100 - 1.3,
      TRUE ~ cpi_1990_100
    ),
    label_y = case_when(
      year == 2000 ~ female_lfpr_25_49_pct + 3.0,
      TRUE ~ female_lfpr_25_49_pct + 2.2
    )
  )

ggplot(
  insight1_path,
  aes(
    x = cpi_1990_100,
    y = female_lfpr_25_49_pct,
    colour = single_female_25_49_pct
  )
) +
  geom_path(aes(group = 1), colour = "grey65", linewidth = 1.0, alpha = 0.85) +
  geom_point(size = 4.8, alpha = 0.95) +
  geom_label(
    data = insight1_path_labels,
    aes(x = label_x, y = label_y, label = year),
    size = 3.5,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  scale_colour_gradient(
    low = "#D8BFD8",
    high = "#6A51A3",
    labels = function(x) paste0(round(x, 1), "%")
  ) +
  scale_x_continuous(
    breaks = seq(100, 220, 20),
    expand = expansion(mult = c(0.05, 0.08))
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    breaks = seq(55, 90, 5),
    expand = expansion(mult = c(0.05, 0.12))
  ) +
  labs(
    title = "Did cost, female LFPR and singlehood move together?",
    subtitle = "Moving right = higher CPI. Moving up = higher female LFPR. Darker points = higher singlehood.",
    x = "Cost of living index, 1990 = 100",
    y = "Female LFPR, ages 25 to 49",
    colour = "Single women,\nages 25 to 49",
    caption = "CPI is indexed to 1990 = 100. LFPR and singlehood are shown in their original percentage units. This shows co-movement, not causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 4. Insight 2 plot A: did births among younger women fall sharply?

```{r}
young_asfr_decade <- asfr_by_age |>
  filter(age_group %in% c("15-19 Years", "20-24 Years", "25-29 Years")) |>
  mutate(
    decade_start = floor(year / 10) * 10,
    decade = paste0(decade_start, "s")
  ) |>
  filter(decade_start >= 1960, decade_start <= 2020) |>
  group_by(decade, decade_start, age_group) |>
  summarise(mean_asfr = mean(asfr, na.rm = TRUE), .groups = "drop") |>
  mutate(
    age_group = factor(age_group, levels = rev(c("15-19 Years", "20-24 Years", "25-29 Years"))),
    label = round(mean_asfr, 0)
  )

ggplot(young_asfr_decade, aes(x = decade, y = age_group, fill = mean_asfr)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(
    aes(label = label),
    size = 4,
    fontface = "bold",
    colour = "grey20"
  ) +
  scale_fill_gradient(
    low = "#FFF7BC",
    high = "#D95F0E",
    labels = scales::label_number()
  ) +
  labs(
    title = "Did births among younger women fall sharply?",
    subtitle = "Average ASFR by decade. Darker cells mean higher fertility intensity.",
    x = "Decade",
    y = "Mother's age group",
    fill = "Mean ASFR",
    caption = "ASFR means age-specific fertility rate, measured as births per 1,000 women in that age group."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )
```

## 5. Insight 2 plot B: did professional work and female education rise as young births fell?

```{r}
asfr_15_19 <- asfr_by_age |>
  filter(age_group == "15-19 Years") |>
  transmute(year, asfr_15_19 = asfr)

asfr_20_24 <- asfr_by_age |>
  filter(age_group == "20-24 Years") |>
  transmute(year, asfr_20_24 = asfr)

insight2_change <- pmet_share |>
  left_join(female_university_share_25_49, by = "year") |>
  left_join(asfr_15_19, by = "year") |>
  left_join(asfr_20_24, by = "year") |>
  filter(year %in% c(2000, 2010, 2020, 2025)) |>
  select(
    year,
    pmet_share_pct,
    female_university_25_49_pct,
    asfr_15_19,
    asfr_20_24
  ) |>
  pivot_longer(
    cols = -year,
    names_to = "measure",
    values_to = "value"
  ) |>
  mutate(
    measure = recode(
      measure,
      pmet_share_pct = "PMET share",
      female_university_25_49_pct = "Female university share",
      asfr_15_19 = "ASFR, age 15 to 19",
      asfr_20_24 = "ASFR, age 20 to 24"
    ),
    direction = if_else(
      measure %in% c("PMET share", "Female university share"),
      "Labour and education increased",
      "Young-age births decreased"
    ),
    value_label = if_else(
      measure %in% c("PMET share", "Female university share"),
      paste0(round(value, 0), "%"),
      paste0(round(value, 0))
    ),
    measure = factor(
      measure,
      levels = rev(c(
        "PMET share",
        "Female university share",
        "ASFR, age 20 to 24",
        "ASFR, age 15 to 19"
      ))
    )
  ) |>
  group_by(measure) |>
  mutate(
    base_value = value[year == 2000],
    change_from_2000 = value - base_value,
    encoded_change = if_else(
      direction == "Young-age births decreased",
      base_value - value,
      value - base_value
    )
  ) |>
  ungroup() |>
  mutate(
    trend_label = case_when(
      year == 2000 ~ "baseline",
      direction == "Labour and education increased" ~ paste0("+", round(change_from_2000, 0)),
      direction == "Young-age births decreased" ~ paste0(round(change_from_2000, 0))
    )
  )

ggplot(insight2_change, aes(x = factor(year), y = measure, fill = encoded_change)) +
  geom_tile(colour = "white", linewidth = 0.8, alpha = 1) +
  geom_text(
    aes(label = paste0(value_label, "\n", trend_label)),
    size = 4,
    fontface = "bold",
    colour = "grey15"
  ) +
  facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
  scale_fill_gradient(
    low = "#DDEAF6",
    high = "#2B8CBE",
    labels = scales::label_number()
  ) +
  labs(
    title = "Did labour and education rise while young-age births fell?",
    subtitle = "Darker cells show larger change from 2000 within each row. Top number = actual value, bottom number = change from 2000.",
    x = "Year",
    y = NULL,
    fill = "Change\nfrom 2000",
    caption = "PMET and education are percentages. ASFR values are births per 1,000 women. Colour intensity compares change within each row, not equivalent units."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text.y = element_text(face = "bold", angle = 0),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )
```

## 6. Insight 2 plot C: did young adulthood shift toward education and professional work?

```{r}
insight2_evidence_board <- pmet_share |>
  left_join(female_university_share_25_49, by = "year") |>
  left_join(asfr_15_19, by = "year") |>
  left_join(asfr_20_24, by = "year") |>
  filter(year >= 2000, year <= 2025) |>
  select(
    year,
    pmet_share_pct,
    female_university_25_49_pct,
    asfr_20_24,
    asfr_15_19
  ) |>
  pivot_longer(
    cols = -year,
    names_to = "measure",
    values_to = "value"
  ) |>
  filter(!is.na(value)) |>
  mutate(
    measure = recode(
      measure,
      pmet_share_pct = "Professional jobs increased",
      female_university_25_49_pct = "Female university attainment increased",
      asfr_20_24 = "Births age 20 to 24 fell",
      asfr_15_19 = "Births age 15 to 19 nearly disappeared"
    ),
    measure = factor(
      measure,
      levels = c(
        "Professional jobs increased",
        "Female university attainment increased",
        "Births age 20 to 24 fell",
        "Births age 15 to 19 nearly disappeared"
      )
    ),
    group = if_else(
      measure %in% c("Professional jobs increased", "Female university attainment increased"),
      "Labour and education",
      "Young-age births"
    )
  )

insight2_evidence_labels <- insight2_evidence_board |>
  group_by(measure) |>
  filter(year %in% range(year)) |>
  ungroup() |>
  mutate(
    label = if_else(
      group == "Labour and education",
      paste0(round(value, 0), "%"),
      paste0(round(value, 0))
    ),
    label_x = if_else(year == min(year), year - 0.7, year + 0.7),
    hjust_label = if_else(year == min(year), 1, 0)
  )

ggplot(insight2_evidence_board, aes(x = year, y = value, fill = group, colour = group)) +
  geom_area(alpha = 0.28, linewidth = 0) +
  geom_line(linewidth = 1.15) +
  geom_point(
    data = insight2_evidence_labels,
    size = 2.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = insight2_evidence_labels,
    aes(x = label_x, label = label, hjust = hjust_label),
    size = 3.4,
    fontface = "bold",
    colour = "grey20",
    show.legend = FALSE
  ) +
  facet_wrap(~ measure, scales = "free_y", ncol = 2) +
  scale_x_continuous(
    breaks = c(2000, 2010, 2020, 2025),
    expand = expansion(mult = c(0.12, 0.16))
  ) +
  scale_fill_manual(
    values = c("Labour and education" = "#8ECAE6", "Young-age births" = "#FDB863")
  ) +
  scale_colour_manual(
    values = c("Labour and education" = "#2B8CBE", "Young-age births" = "#D95F0E")
  ) +
  labs(
    title = "Did young adulthood shift toward education and professional work?",
    subtitle = "Each panel keeps its own unit. The top pattern rises while young-age ASFR falls.",
    x = "Year",
    y = NULL,
    fill = NULL,
    colour = NULL,
    caption = "PMET and education are percentages. ASFR values are births per 1,000 women."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", hjust = 0),
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 7. Insight 3 plot A: did marriage and first birth shift later?

```{r}
family_timing <- bind_rows(
  resident_brides_first_marriage |>
    filter(year >= 1990, year <= 2024) |>
    transmute(
      year,
      milestone = "First marriage",
      median_age = median_age_first_marriage
    ),
  resident_first_birth_age |>
    filter(year >= 1990, year <= 2024) |>
    transmute(
      year,
      milestone = "First birth",
      median_age = median_age_first_birth
    )
) |>
  mutate(
    milestone = factor(
      milestone,
      levels = c("First marriage", "First birth")
    )
  )

family_timing_labels <- family_timing |>
  group_by(milestone) |>
  filter(year %in% c(min(year), max(year))) |>
  ungroup() |>
  mutate(
    label = paste0(year, ": ", round(median_age, 1)),
    label_x = case_when(
      year == min(year) ~ year - 0.8,
      TRUE ~ year + 0.45
    ),
    label_y = case_when(
      milestone == "First marriage" & year == min(year) ~ median_age - 0.25,
      milestone == "First birth" & year == min(year) ~ median_age + 0.25,
      milestone == "First marriage" ~ median_age - 0.18,
      TRUE ~ median_age + 0.18
    ),
    hjust_label = if_else(year == min(year), 1, 0)
  )

family_timing_change <- family_timing |>
  group_by(milestone) |>
  summarise(
    start_age = median_age[year == min(year)],
    end_age = median_age[year == max(year)],
    change_years = end_age - start_age,
    .groups = "drop"
  ) |>
  mutate(
    label = paste0(milestone, " shifted +", round(change_years, 1), " years")
  )

family_timing_note <- family_timing_change |>
  summarise(note = paste(label, collapse = "\n")) |>
  mutate(
    year = 1991,
    median_age = 32.3
  )

ggplot(family_timing, aes(x = year, y = median_age, colour = milestone)) +
  geom_line(linewidth = 1.25) +
  geom_point(
    data = family_timing_labels,
    size = 3.4,
    show.legend = FALSE
  ) +
  geom_label(
    data = family_timing_labels,
    aes(x = label_x, y = label_y, label = label, hjust = hjust_label),
    size = 3.4,
    fontface = "bold",
    label.size = 0.18,
    fill = "white",
    show.legend = FALSE
  ) +
  geom_text(
    data = family_timing_note,
    aes(x = year, y = median_age, label = note),
    inherit.aes = FALSE,
    hjust = 0,
    size = 3.5,
    fontface = "bold",
    colour = "grey25"
  ) +
  scale_colour_manual(
    values = c(
      "First marriage" = "#2F4B7C",
      "First birth" = "#7A5195"
    )
  ) +
  scale_x_continuous(
    breaks = seq(1990, 2025, 5),
    expand = expansion(mult = c(0.09, 0.16))
  ) +
  scale_y_continuous(
    breaks = seq(25, 33, 1),
    limits = c(24.7, 32.7)
  ) +
  labs(
    title = "Did marriage and first birth shift later?",
    subtitle = "Median ages for resident brides and resident births, 1990 to 2024.",
    x = "Year",
    y = "Median age",
    colour = NULL,
    caption = "The two medians describe population timing. They do not track the same individuals."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## 8. Insight 3 plot B: did completed family size shift smaller?

```{r}
family_size_selected <- family_size_40_49 |>
  mutate(
    family_size_group = case_when(
      children_born == "None" ~ "None",
      children_born == "1 Child" ~ "1 child",
      children_born == "2 Children" ~ "2 children",
      children_born %in% c("3 Children", "4 Or More Children") ~ "3 or more children",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(family_size_group), year %in% c(1990, 2000, 2010, 2020, 2025)) |>
  group_by(year, family_size_group) |>
  summarise(share_pct = sum(share_pct, na.rm = TRUE), .groups = "drop") |>
  mutate(
    family_size_group = factor(
      family_size_group,
      levels = c("3 or more children", "2 children", "1 child", "None")
    ),
    label = if_else(share_pct >= 7, paste0(round(share_pct, 0), "%"), "")
  )

ggplot(family_size_selected, aes(x = factor(year), y = share_pct, fill = family_size_group)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 3.6,
    fontface = "bold",
    colour = "grey20"
  ) +
  scale_y_continuous(
    labels = scales::label_percent(scale = 1),
    breaks = seq(0, 100, 25),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c(
      "3 or more children" = "#C94B55",
      "2 children" = "#F2D98D",
      "1 child" = "#74C69D",
      "None" = "#6B7280"
    )
  ) +
  labs(
    title = "Did completed family size shift smaller?",
    subtitle = "Share of ever-married resident females aged 40 to 49 by number of children born.",
    x = "Year",
    y = "Share of women aged 40 to 49",
    fill = "Children born",
    caption = "Selected benchmark years are shown to keep the family-size shift readable."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 9. Insight 4 plot: did recessions and policies explain fertility changes?

```{r}
policy_events <- tribble(
  ~event_year, ~event_label, ~policy_stance,
  1972, "Stop at Two", "Anti-natalist",
  1987, "Have Three or More", "Pro-natalist",
  2001, "Baby Bonus", "Pro-natalist",
  2013, "Marriage and Parenthood package", "Pro-natalist"
)

economic_shocks <- tribble(
  ~shock_year, ~shock_label, ~severity, ~quarters_below_peak,
  1985, "1985", "Severe", 6,
  1998, "1998", "Severe", 6,
  2001, "2001", "Moderate", 5,
  2009, "2009", "Most severe", 6,
  2020, "2020", "Most severe", 4
)

policy_window <- policy_events |>
  crossing(relative_year = -5:10) |>
  mutate(year = event_year + relative_year) |>
  left_join(tfr_annual, by = "year") |>
  mutate(
    event_label = factor(
      event_label,
      levels = c("Stop at Two", "Have Three or More", "Baby Bonus", "Marriage and Parenthood package")
    )
  )

shock_window <- policy_events |>
  crossing(economic_shocks) |>
  mutate(
    relative_year = shock_year - event_year,
    recovery_years = quarters_below_peak / 4,
    relative_recovery_end = relative_year + recovery_years,
    event_label = factor(
      event_label,
      levels = c("Stop at Two", "Have Three or More", "Baby Bonus", "Marriage and Parenthood package")
    )
  ) |>
  filter(relative_year <= 10, relative_recovery_end >= -5) |>
  mutate(
    xmin = pmax(relative_year, -5),
    xmax = pmin(relative_recovery_end, 10)
  )

ggplot(policy_window, aes(x = relative_year, y = tfr, colour = policy_stance)) +
  geom_rect(
    data = shock_window,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = -Inf,
      ymax = Inf,
      fill = severity
    ),
    inherit.aes = FALSE,
    alpha = 0.24
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey35") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.3) +
  geom_label(
    data = shock_window,
    aes(x = (xmin + xmax) / 2, y = Inf, label = shock_label),
    inherit.aes = FALSE,
    vjust = 1.2,
    size = 2.8,
    label.size = 0.15,
    fill = "white",
    colour = "grey35"
  ) +
  facet_wrap(~ event_label, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("Anti-natalist" = "#D95F02", "Pro-natalist" = "#1B9E77")) +
  scale_fill_manual(
    values = c(
      "Most severe" = "#B2182B",
      "Severe" = "#8073AC",
      "Moderate" = "#D8DAEB"
    )
  ) +
  scale_x_continuous(breaks = c(-5, 0, 5, 10)) +
  labs(
    title = "Did recessions and policies explain fertility changes?",
    subtitle = "Dashed line = policy year. Shaded bands show economic recovery duration, with warmer colours indicating higher severity.",
    x = "Years from policy",
    y = "TFR",
    colour = "Policy stance",
    fill = "Economic shock severity",
    caption = "Policy and recession markers provide context only. Timing is not causal proof."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )
```
